Imputation of Incomplete Multilevel Data with \pkg{mice}

### 2022-12-13 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd

[TODO: Add environment info, seed and version number(s) somewhere.] Set up the R environment and load the necessary packages:
```{r env, message=FALSE, warning=FALSE}
set.seed(2022)
library(mice)         # for imputation
library(miceadds)     # for imputation
library(ggmice)       # for visualization
library(ggplot2)      # for visualization
library(dplyr)        # for data wrangling
library(lme4)         # for multilevel modeling
library(mitml)        # for multilevel pooling
```
<!-- library(broom.mixed)  # for multilevel estimates -->

@@

To explore the missingness, it is wise to look at the missing data pattern:
```{r pattern, fig.height=7.1}
plot_pattern(dat, rotate = TRUE)  # plot missingness pattern
```

@@

Load the data:
```{r}
data("hiv",  package = "GJRM")
# We select  5 predictor variables over 9 regions
colnames(hiv)
hivdata <- hiv[,c("hiv","hivconsent","age","marital","condom","highhiv","interviewerID","region","language")]
```

Recode variables:
```{r}
# Study/group variable has to be recoded as integer
hivdata$region<-as.integer(hivdata$region)

# Categorical variables have to be recoded as factor
# to use the 2l.binary imputation method, it is required 
# that the level names should not have any space between words

hivdata$hiv <- as.factor(hivdata$hiv) #to use the 2l.heckman method, it is required that
#the missing variable is stored as a factor in the dataset, otherwise the method will 
#apply the imputation correction for a  missing continuous variable instead for 
#a missing binary variable,which is in this case the binary response of the hiv test.

hivdata$marital <- as.factor(hivdata$marital)
levels(hivdata$marital)<-c("never_married","currently_married","formerly_married")
hivdata$condom <- as.factor(hivdata$condom)
levels(hivdata$condom)<-c("No_Condom_Last_Intercourse","Condom_Last_Intercourse")
hivdata$highhiv <- as.factor(hivdata$highhiv)
levels(hivdata$highhiv)<-c("Not_High_Risk_of_HIV","High_Risk_of_HIV")
hivdata$interviewerID <- as.factor(hivdata$interviewerID)

hivdata$interviewerID <-  as.factor(as.character(hivdata$interviewerID))
interv<-as.data.frame(table(hivdata$interviewerID,hivdata$region))
```

We obtain here the random effects for each interviewer, this is an approximation of the interviewer's skill which will be used as an exclusion constraint. Here, since the location of the interviewer was not randomly assigned to the subjects, the assignment was corrected for region and language.
```{r}
hivdata$hivconsent <- as.factor(hivdata$hivconsent)
ID_mixed <- lme4::glmer(hivconsent ~ region + language+(1 | interviewerID), data = hivdata, family = binomial)
reffect  <- ranef(ID_mixed)$interviewerID
reffect$interviewerID <- levels(hivdata$interviewerID)
colnames(reffect)     <- c("IDreffect","interviewerID")
hivdata  <-  merge(hivdata, reffect, by="interviewerID", all.x=TRUE)

hivdata$interviewerID<-NULL
hivdata$hivconsent<-NULL
hivdata$language<-NULL
```

Set the Heckman model as imputation method
```{r}
#Set prediction matrix and methods
ini <- mice(hivdata, maxit = 0)
meth<-ini$method
meth["hiv"]<-"2l.binary"
pred <- ini$pred
pred[,"region"] <- 0
pred["region",] <- 0
pred["hiv","region"]<- -2
pred["hiv","IDreffect"]<- 0

# Heckman model
pred["hiv","IDreffect"] <- -3
meth<-ini$method
meth["hiv"]<-"2l.heckman"
```

Impute the data:
```{r eval=FALSE}
imp <- mice( hivdata, # dataset with missing values
                        m = 10,   # number of imputations
                        maxit = 1,
                        seed = 1234, #seed attached to the dataID
                        meth = meth, #imputation method vector
                        pred = pred, #imputation predictors matrix
                        print = T,
                        meta_method="reml",
                        pmm=FALSE)
```

>>>>

[TODO: Add environment info, seed and version number(s) somewhere.] Set up the R environment and load the necessary packages:
```{r env, message=FALSE, warning=FALSE}
set.seed(2022)
library(mice)         # for imputation
library(miceadds)     # for imputation
library(ggmice)       # for visualization
library(ggplot2)      # for visualization
library(dplyr)        # for data wrangling
library(lme4)         # for multilevel modeling
library(mitml)        # for multilevel pooling
library(miceheckman)  # for imputation
```
<!-- library(broom.mixed)  # for multilevel estimates -->

@@

To explore the missingness, it is wise to look at the missing data pattern:
```{r pattern, fig.height=7.1}
plot_pattern(dat, rotate = TRUE, npat = 10L)  # plot missingness pattern
```

@@

Load the data:
```{r}
data("obesity",  package = "miceheckman")
```

We obtain here the random effects for each interviewer, this is an approximation of the interviewer's skill which will be used as an exclusion constraint. Here, since the location of the interviewer was not randomly assigned to the subjects, the assignment was corrected for region and language.

Set the Heckman model as imputation method
```{r}
#Set prediction matrix and methods
ini <- mice(obesity, maxit = 0)
meth<-ini$method
meth["weight"]<-"2l.heckman"
pred <- ini$pred
pred["weight","cluster"]<- -2
pred["weight","rt"] <- -3
```

Impute the data:
```{r eval=FALSE}
imp <- mice(obesity, # dataset with missing values
                        m = 10,   # number of imputations
                        maxit = 1,
                        seed = 1234, #seed attached to the dataID
                        meth = meth, #imputation method vector
                        pred = pred, #imputation predictors matrix
                        print = T,
                        meta_method="reml",
                        pmm=FALSE)
```
